---
title: "In class Exercise 5"
execute: 
  warning: false
date: "`r Sys.Date()`"
highlight-style: dracula
---

## **1. Installing R Packages**

```{r}
pacman::p_load(sf, sfdep, tmap, tidyverse)
```

## **2. The Data**

-   Hunan geospatial data in ESRI shapefile format
-   Hunan_2012 attribute data in csv format

### Importing geospatial data

```{r}
hunan <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan")
```

### Importing aspatial data

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

```{r}
hunan_GDPPC <- read_csv("data/aspatial/Hunan_GDPPC.csv")
```

### Performing relational join

```{r}
hunan_GDPPC <- left_join(hunan,hunan2012) %>%
  select(1:4, 7, 15)
```

### Plotting a choropleth map

```{r}
tmap_mode("plot")
tm_shape(hunan_GDPPC) +
  tm_fill("GDPPC",
          n = 5,
          style = "quantile",
          palette = "Greens") +
  tm_borders(alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar() +
  tm_layout(main.title = "Equal quantile classification",
            main.title.size = 1,
            main.title.position = "center",
            frame = TRUE,
            legend.height = 0.45,
            legend.width = 0.35) + 
  tm_grid(alpha = 0.2)
```

## **3. Global Measures of Spatial Association**

### Derive contiguity weights with Queen's method

```{r}
wm_q <- hunan_GDPPC %>%
  mutate(nb = st_contiguity(geometry),
         wt = st_weights(nb,
                         style="W"),
         .before = 1)
```

### Computing Global Moran's I

```{r}
moranI <- global_moran(wm_q$GDPPC,
                       wm_q$nb,
                       wm_q$wt)

glimpse(moranI)
```

### Performing Global Moran's I permutation test

```{r}
set.seed(1234)
```

```{r}
global_moran_perm(wm_q$GDPPC,
                       wm_q$nb,
                       wm_q$wt,
                  nsim = 99)
```

P-value is smaller than alpha value of 0.05. Hence, we have enough statistical evidence to reject the null hypothesis.